Quasars Probing Quasars IX. The Kinematics of the Circumgalactic Medium Surrounding $z\sim2$ Quasars [v1.4]

### 2017-09-28 QPQ9/Submit/qpq9_submit.tex

\begin{abstract}
We examine the kinematics of the gas in the environments of galaxies hosting quasars at $z\sim2$. 
We employ 148 projected quasar pairs to study the circumgalactic gas of the foreground quasars in 
absorption. The sample selects foreground quasars with precise redshift measurements, using 
emission-lines with precision $\lesssim300\,{\rm km\,s^{-1}}$ and average offsets from the 
systemic redshift $\lesssim|100\,{\rm km\,s^{-1}}|$. We stack the background quasar spectra at the 
foreground quasar's systemic redshift to study the mean absorption in \ion{C}{2}, \ion{C}{4}, 
and \ion{Mg}{2}. We find that the mean absorptions exhibit large velocity widths 
$\sigma_v>300\,{\rm km\,s^{-1}}$. The observed widths are consistent with gas in gravitational 
motion and Hubble flow, and galactic-scale outflows are not required to explain the large 
widths. Furthermore, we find that the mean absorptions are asymmetric about the systemic redshift. 
The mean absorption centroids exhibit small redshift relative to the 
systemic $\delta v\approx+200\,{\rm km\,s^{-1}}$, with large intrinsic scatter in the centroid 
velocities of the individual absorption systems. We show that the observed offsets may be produced 
if (i) the ionizing radiation from the foreground quasars is anisotropic or intermittent; (ii) the 
gas is not flowing into the galaxy.  
\end{abstract}

@@

In this paper (hereafter QPQ9), we examine the flows of gas in the environments of massive 
galaxies hosting quasars. Our approach leverages a large dataset of quasar pairs 
\citep[][hereafter QPQ1]{QPQ1} to use the standard techniques of absorption-line spectroscopy 
with background quasars. These quasar pairs 
have angular separations that correspond to less than 300\,kpc projected separation at the 
foreground quasar's redshift. Our previous publications from these quasar pairs have established 
that these galaxies are surrounded by a massive, cool, and enriched CGM 
\citep[QPQ5, QPQ6, QPQ7:][]{QPQ5,QPQ6,QPQ7}. We have collected a sample of 102 background 
spectra that are paired with foreground quasars with precisely measured redshifts. Among the
sightlines in the QPQ9 sample, 13 have spectral resolution $R>5000$ from echellette or echelle
observations, and have been analyzed separately in QPQ8 (see their Figure 8, 9, and 10, and their 
Appendix). In QPQ8, where the individual metal-bearing absorption components are resolved, 15 out 
of the 21 components are at positive velocities relative to the systemic redshift. For this
current work, we stack spectra of all resolutions instead of performing a component-by-component
analysis. Our primary scientific interests are twofold: (i) search for signatures of galactic-scale
outflows from the central galaxy, presumably driven by recent star formation and/or active galactic
nuclei feedback; (ii) characterize the dynamics of the gas around these massive systems. We further
describe an aspect of this experiment that offers a unique opportunity to study galactic-scale
flows: we argue that, the anisotropic or intermittent radiation from the foreground quasars may
break the symmetry in the velocity field of circumgalactic absorbers. If the ionizing radiation
field is asymmetric, the absorbers may also distribute asymmetrically. Alternatively, finite quasar
lifetime will result in different radiation fields impinged on the gas closer to versus further
away from the background quasar, due to different light travel times. \cite{KirkmanTytler08}
reported an asymmetry in \ion{H}{1} absorption on scales larger than the CGM, and gave similar
arguments on anisotropy or intermittence.

@@

We further add to the QPQ dataset with quasar pairs selected from the public dataset of 
igmspec\footnote{https://github.com/specdb/igmspec} \citep{Prochaska+17}, which includes the 
spectra from the quasar catalogs based upon the Sloan Digital Sky Survey Seventh Data Release 
\citep{Schneider+10} and the Twelfth Data Release \citep{Paris+17}. We only select pairs with 
$z_{\rm fg}$ measurable using a robust \ion{Mg}{2}\,2800 emission-line. 
We reach a final sample size of 102. 
Figure~\ref{fig:experiment} summarizes the experimental design. We refer the reader to previous 
QPQ publications for the details on the emission-line centering algorithm, data reduction, and
continuum normalization (QPQ1; QPQ6; QPQ8).

@@

\begin{figure}
%\includegraphics[width=3.5in]{Figures/fig_stacks_and_fits.pdf}
\includegraphics[width=3.5in]{f02.pdf}
\caption{Mean and median absorption centered at \ion{C}{2}\,1334, \ion{C}{4}\,1548, and 
\ion{Mg}{2}\,2796 of the foreground quasars for all QPQ9 pairs. The composites are shown in thin, 
black. Overplotted on the composites are Gaussian fits, normalized to pseudo-continua far away 
from a velocity of $0\,{\rm km\,s^{-1}}$ relative to $z_{\rm fg}$, and are shown in thick, blue. 
For the doublets, a second Gaussian with a fixed mean separation and a tied standard deviation is 
included in the modeling. The absorptions frequently exhibit large velocity widths. The dashed 
lines mark the centroids, which show small positive velocity offsets from $z_{\rm fg}$. The 
$1\sigma$ modeling error for the centroid and dispersion of the \ion{C}{2} mean stack are 
$25\,{\rm km\,s^{-1}}$ and $27\,{\rm km\,s^{-1}}$ respectively, which is also the typical modeling
errors of the other stacks.
}
\label{fig:stacks_and_fits}
\end{figure}

@@

Given the small sample size, one must scrutinize the statistical significance of the measured 
velocity offsets. To assess the statistical significance, we perform a boostrap analysis by 
randomly resampling from the full sample 10000 times. We introduce a Gaussian absorption profile 
to model each bootstrap realization, and quote the standard deviation in the bootstrap 
realizations to be the scatter in the centroids of the data. The scatters 
$\approx100\,{\rm km\,s^{-1}}$ are comparable to the measured offsets, indicating large intrinsic 
variation in quasar CGM environments (see also QPQ8). In Figure~\ref{fig:histogram_cen}, we show 
the distribution of the absorption centroids from bootstrapping on the \ion{C}{2} mean stack. We 
find that 97\% of the centroids are at positive velocities. We place a generous $3\sigma$ upper
limit to the small offset of the centroid from $z_{\rm fg}$ for the \ion{C}{2} mean absorption at 
$\delta v<+526\,{\rm km\,s^{-1}}$. Given the large intrinsic scatter, we do not attempt to explore
whether there exists relative asymmetry among the \ion{C}{2}, \ion{C}{4}, and \ion{Mg}{2} 
absorption.

@@

\begin{figure}
%\includegraphics[width=2.5in]{Figures/fig_contour.pdf}
\includegraphics[width=2.5in]{f06.pdf}
\caption{
Probability distributions of the parameters $W_{\rm CII}$ and $M_{\rm halo}$. The plot shows the 
degeneracy between $W_{\rm CII}$ and $M_{\rm halo}$ in recovering the intrinsic width of the 
absorption profile. We mark contours for points that produce an absorption 
profile of width that is 1, 2, and 3 times the modeling error away from the observed intrinsic 
width. The intrinsic velocity width corresponding to typical QPQ halo mass is marked with a plus 
sign, and is well contained within the 1$\sigma$ contour.}
\label{fig:contour}
\end{figure}

@@

Since \ion{C}{2} systems arise in optically thick absorbers, we may adopt the clustering analysis 
results of QPQ6. In the absence of clustering, the expected number of absorbers per unit redshift 
interval for Lyman limit systems, super Lyman limit systems, and damped Ly$\alpha$ systems are 
respectively $\ell_{\rm IGM}^{\rm LLS}(z)\approx1.05((1+z)/(1+2.5))^{2.1}$, 
$\ell_{\rm IGM}^{\rm SLLS}(z)\approx0.44((1+z)/(1+2.5))^{2.1}$, and 
$\ell_{\rm IGM}^{\rm DLA}(z)\approx0.2((1+z)/(1+2.5))^{2.1}$. The quasar-absorber correlation 
functions for Lyman limit systems, super Lyman limit systems, and damped Ly$\alpha$ systems are 
respectively $\xi_{\rm QA}^{\rm LLS}(r)=(r/(12.5\,h^{-1}\,{\rm Mpc}))^{-1.68}$, 
$\xi_{\rm QA}^{\rm SLLS}(r)=(r/(14.0\,h^{-1}\,{\rm Mpc}))^{-1.68}$, and 
$\xi_{\rm QA}^{\rm DLA}(r)=(r/(3.9\,h^{-1}\,{\rm Mpc}))^{-1.6}$. 
For each quasar pair, we calculate the expected number of optically thick absorbers within 
$\pm3000\,{\rm km\,s^{-1}}$ at a distance $R_\perp$ from the foreground quasar and at 
$z_{\rm fg}$. Then we generate 1000 mock sightlines. The number of absorbers for each mock 
spectrum is randomly selected from a Poisson distribution with mean equal to the expected number 
calculated as above. The absorbers are randomly assigned Hubble velocities, with a probability 
distribution according to the quasar-absorber correlation functions. The absorbers are randomly 
assigned additional peculiar velocities drawn from a normal distribution with mean equal to 
$0\,{\rm km\,s^{-1}}$ and scatter equal to $\sigma_{\rm rms}$. For each absorber, we assume a rest
equivalent width for \ion{C}{2} $W_{\rm CII}$ and a Gaussian absorption profile. We repeat the 
above procedure for all 33 quasar pairs, and create a mean stack of the 40000 mock spectra
generated. We fit a Gaussian absorption profile multiplied to a constant continuum level to model 
the stack of mock spectra. We adjust the $W_{\rm CII}$ adopted for the absorbers until the 
amplitude of the best-fit Gaussian of the stack of mock spectra matches the amplitude of the stack
of the observational data. We find that $W_{\rm CII}\approx0.5\,{\rm \AA}$ well reproduces the
amplitude, and the dispersion of the Gaussian absorption model is insensitive to the assumed
$W_{\rm CII}$ or line profile for one absorber.

>>>>

\begin{abstract}
We examine the kinematics of the gas in the environments of galaxies hosting quasars at $z\sim2$. 
We employ 148 projected quasar pairs to study the circumgalactic gas of the foreground quasars in
absorption. The sample selects foreground quasars with precise redshift measurements, using 
emission-lines with precision $\lesssim300\,{\rm km\,s^{-1}}$ and average offsets from the 
systemic redshift $\lesssim|100\,{\rm km\,s^{-1}}|$. We stack the background quasar spectra at the 
foreground quasar's systemic redshift to study the mean absorption in \ion{C}{2}, \ion{C}{4}, 
and \ion{Mg}{2}. We find that the mean absorptions exhibit large velocity widths 
$\sigma_v>300\,{\rm km\,s^{-1}}$. The observed widths are consistent with gas in gravitational 
motion and Hubble flow, and galactic-scale outflows are not be required to explain the large
widths. Furthermore, we find that the mean absorptions are asymmetric about the systemic redshift. 
The mean absorption centroids exhibit small redshift relative to the 
systemic $\delta v\approx+200\,{\rm km\,s^{-1}}$, with large intrinsic scatter in the centroid 
velocities of the individual absorption systems. We show that the observed offsets may be produced 
if (i) the ionizing radiation from the foreground quasars is anisotropic or intermittent; (ii) the 
gas is not flowing into the galaxy.  
\end{abstract}

@@

In this paper (hereafter QPQ9), we examine the flows of gas in the environments of massive 
galaxies hosting quasars. Our approach leverages a large dataset of quasar pairs 
\citep[][hereafter QPQ1]{QPQ1} to use the standard techniques of absorption-line spectroscopy 
with background quasars. These quasar pairs 
have angular separations that correspond to less than 300\,kpc projected separation at the 
foreground quasar's redshift. Our previous publications from these quasar pairs have established 
that these galaxies are surrounded by a massive, cool, and enriched CGM 
\citep[QPQ5, QPQ6, QPQ7:][]{QPQ5,QPQ6,QPQ7}. We have collected a sample of 148 background 
spectra that are paired with foreground quasars with precisely measured redshifts. Among the
sightlines in the QPQ9 sample, 13 have spectral resolution $R>5000$ from echellette or echelle
observations, and have been analyzed separately in QPQ8 (see their Figure 8, 9, and 10, and their 
Appendix). In QPQ8, where the individual metal-bearing absorption components are resolved, 15 out 
of the 21 components are at positive velocities relative to the systemic redshift. For this
current work, we stack spectra of all resolutions instead of performing a component-by-component
analysis. Our primary scientific interests are twofold: (i) search for signatures of galactic-scale
outflows from the central galaxy, presumably driven by recent star formation and/or active galactic
nuclei feedback; (ii) characterize the dynamics of the gas around these massive systems. We further
describe an aspect of this experiment that offers a unique opportunity to study galactic-scale
flows: we argue that, the anisotropic or intermittent radiation from the foreground quasars may
break the symmetry in the velocity field of circumgalactic absorbers. If the ionizing radiation
field is asymmetric, the absorbers may also distribute asymmetrically. Alternatively, finite quasar
lifetime will result in different radiation fields impinged on the gas closer to versus further
away from the background quasar, due to different light travel times. \cite{KirkmanTytler08}
reported an asymmetry in \ion{H}{1} absorption on scales larger than the CGM, and gave similar
arguments on anisotropy or intermittence.

@@

We further add to the QPQ dataset with quasar pairs selected from the public dataset of 
igmspec\footnote{https://github.com/specdb/igmspec} \citep{Prochaska+17}, which includes the 
spectra from the quasar catalogs based upon the Sloan Digital Sky Survey Seventh Data Release 
\citep{Schneider+10} and the Twelfth Data Release \citep{Paris+17}. We only select pairs with 
$z_{\rm fg}$ measurable using a robust \ion{Mg}{2}\,2800 emission-line. 
We reach a final sample size of 148. 
Figure~\ref{fig:experiment} summarizes the experimental design. We refer the reader to previous 
QPQ publications for the details on the emission-line centering algorithm, data reduction, and
continuum normalization (QPQ1; QPQ6; QPQ8).

@@

\begin{figure}
%\includegraphics[width=3.5in]{Figures/fig_stacks_and_fits.pdf}
\includegraphics[width=3.5in]{f02.pdf}
\caption{Mean and median absorption centered at \ion{C}{2}\,1334, \ion{C}{4}\,1548, and 
\ion{Mg}{2}\,2796 of the foreground quasars for all QPQ9 pairs. The composites are shown in thin, 
black. Overplotted on the composites are Gaussian fits, normalized to pseudo-continua far away 
from a velocity of $0\,{\rm km\,s^{-1}}$ relative to $z_{\rm fg}$, and are shown in thick, blue. 
For the doublets, a second Gaussian with a fixed mean separation and a tied standard deviation is 
included in the modeling. The absorptions frequently exhibit large velocity widths. The blue dashed 
lines mark the centroids, which show small positive velocity offsets from $z_{\rm fg}$. The 
$1\sigma$ modeling error for the centroid and dispersion of the \ion{C}{2} mean stack are 
$25\,{\rm km\,s^{-1}}$ and $27\,{\rm km\,s^{-1}}$ respectively, which is also the typical modeling
errors of the other stacks.
}
\label{fig:stacks_and_fits}
\end{figure}

@@

To assess the statistical significance of the measured velocity offsets, we perform a boostrap 
analysis by randomly resampling from the full sample 10000 times. We introduce a Gaussian 
absorption profile to model each bootstrap realization, and quote the standard deviation in the 
bootstrap realizations to be the scatter in the centroids of the data. The scatters 
$\approx100\,{\rm km\,s^{-1}}$ are comparable to the measured offsets, indicating large intrinsic 
variation in quasar CGM environments (see also QPQ8). In Figure~\ref{fig:histogram_cen}, we show 
the distribution of the absorption centroids from bootstrapping on the \ion{C}{2} mean stack. We 
find that 97\% of the centroids are at positive velocities. We place a generous $3\sigma$ upper
limit to the small offset of the centroid from $z_{\rm fg}$ for the \ion{C}{2} mean absorption at 
$\delta v<+526\,{\rm km\,s^{-1}}$. Given the large intrinsic scatter, we do not attempt to explore
whether there exists relative asymmetry among the \ion{C}{2}, \ion{C}{4}, and \ion{Mg}{2} 
absorption.

@@

\begin{figure}
%\includegraphics[width=2.5in]{Figures/fig_contour.pdf}
\includegraphics[width=2.5in]{f06.pdf}
\caption{
Probability distributions of the parameters $W_{\rm CII}$ and $M_{\rm halo}$. The plot shows the 
degeneracy between $W_{\rm CII}$ and $M_{\rm halo}$ in recovering the intrinsic width of the 
absorption profile. We mark contours for points that produce an absorption 
profile of width that is 1, 2, and 3 times the modeling error away from the observed intrinsic 
width. The intrinsic velocity width corresponding to typical QPQ halo mass is marked with a plus 
sign, and is contained within the 2$\sigma$ contour.}
\label{fig:contour}
\end{figure}

@@

Since \ion{C}{2} systems arise in optically thick absorbers, we may adopt the clustering analysis 
results of QPQ6. In the absence of clustering, the expected number of absorbers per unit redshift 
interval for Lyman limit systems, super Lyman limit systems, and damped Ly$\alpha$ systems are 
respectively $\ell_{\rm IGM}^{\rm LLS}(z)\approx1.05((1+z)/(1+2.5))^{2.1}$, 
$\ell_{\rm IGM}^{\rm SLLS}(z)\approx0.44((1+z)/(1+2.5))^{2.1}$, and 
$\ell_{\rm IGM}^{\rm DLA}(z)\approx0.2((1+z)/(1+2.5))^{2.1}$. The quasar-absorber correlation 
functions for Lyman limit systems, super Lyman limit systems, and damped Ly$\alpha$ systems are 
respectively $\xi_{\rm QA}^{\rm LLS}(r)=(r/(12.5\,h^{-1}\,{\rm Mpc}))^{-1.68}$, 
$\xi_{\rm QA}^{\rm SLLS}(r)=(r/(14.0\,h^{-1}\,{\rm Mpc}))^{-1.68}$, and 
$\xi_{\rm QA}^{\rm DLA}(r)=(r/(3.9\,h^{-1}\,{\rm Mpc}))^{-1.6}$. 
For each quasar pair, we calculate the expected number of optically thick absorbers within 
$\pm3000\,{\rm km\,s^{-1}}$ at a distance $R_\perp$ from the foreground quasar and at 
$z_{\rm fg}$. Then we generate 1000 mock sightlines. The number of absorbers for each mock 
spectrum is randomly selected from a Poisson distribution with mean equal to the expected number 
calculated as above. The absorbers are randomly assigned Hubble velocities, with a probability 
distribution according to the quasar-absorber correlation functions. The absorbers are randomly 
assigned additional peculiar velocities drawn from a normal distribution with mean equal to 
$0\,{\rm km\,s^{-1}}$ and scatter equal to $\sigma_{\rm rms}$. For each absorber, we assume a rest
equivalent width for \ion{C}{2} $W_{\rm CII}$ and a Gaussian absorption profile. We repeat the 
above procedure for all 40 quasar pairs, and create a mean stack of the 40000 mock spectra
generated. We fit a Gaussian absorption profile multiplied to a constant continuum level to model 
the stack of mock spectra. We adjust the $W_{\rm CII}$ adopted for the absorbers until the 
amplitude of the best-fit Gaussian of the stack of mock spectra matches the amplitude of the stack
of the observational data. We find that $W_{\rm CII}\approx0.5\,{\rm \AA}$ well reproduces the
amplitude, and the dispersion of the Gaussian absorption model is insensitive to the assumed
$W_{\rm CII}$ or line profile for one absorber.

### 2017-09-28 QPQ9/qpq9_draft.tex

\cite{Eftekharzadeh+15} measured the clustering of quasars in the range $2.2<z<2.8$ while 
\cite{RodriguezTorres+16} measured the clustering of quasars in the range $1.8<z<2.2$. They 
estimated that these quasars are hosted by dark matter halos with mass 
$M_{\rm halo}=10^{12.5}\,M_\odot$ and $M_{\rm halo}10^{12.6}\,M_\odot$ respectively. If dark 
matter halos hosting QPQ9 quasars have a characteristic mass $10^{12.6}\,M_\odot$ and follow an 
NFW profile \citep{NavarroFrenkWhite97} with concentration 
parameter $c=4$, at $z\approx2$ the maximum circular velocity is $345\,{\rm km\,s^{-1}}$. 
\cite{TormenBouchetWhite97} found that the maximum circular velocity is $\approx1.4$ times the 
maximum of the one-dimensional velocity dispersion $\sigma_{\rm 1D}$. Hence, the line-of-sight 
velocity dispersion typical of QPQ9 halos is $\sigma_{\rm 1D}=246\,{\rm km\,s^{-1}}$. We can 
investigate whether gravitational motions and Hubble flows are sufficient to reproduce the 
dispersion in the data, using Monte Carlo methods to simulate the absorption signals.

@@

JXP and MWL acknowledge support from the National Science Foundation (NSF) grants AST-1412981. The 
authors gratefully acknowledge the support which enabled these observations at the Keck, Gemini, 
Large Binocular Telescope, Very Large Telescope, Las Campanas, and Palomar Observatories. MWL 
thanks T.-K. Chan for a discussion on CGM simulations and Chi Po Choi for a discussion on 
statistics.

>>>>

\cite{Eftekharzadeh+15} measured the clustering of quasars in the range $2.2<z<2.8$ while 
\cite{RodriguezTorres+17} measured the clustering of quasars in the range $1.8<z<2.2$. They 
estimated that these quasars are hosted by dark matter halos with mass 
$M_{\rm halo}=10^{12.5}\,M_\odot$ and $M_{\rm halo}10^{12.6}\,M_\odot$ respectively. If dark 
matter halos hosting QPQ9 quasars have a characteristic mass $10^{12.6}\,M_\odot$ and follow an 
NFW profile \citep{NavarroFrenkWhite97} with concentration 
parameter $c=4$, at $z\approx2$ the maximum circular velocity is $345\,{\rm km\,s^{-1}}$. 
\cite{TormenBouchetWhite97} found that the maximum circular velocity is $\approx1.4$ times the 
maximum of the one-dimensional velocity dispersion $\sigma_{\rm 1D}$. Hence, the line-of-sight 
velocity dispersion typical of QPQ9 halos is $\sigma_{\rm 1D}=246\,{\rm km\,s^{-1}}$. We can 
investigate whether gravitational motions and Hubble flows are sufficient to reproduce the 
dispersion in the data, using Monte Carlo methods to simulate the absorption signals.

@@

JXP and MWL acknowledge support from the National Science Foundation (NSF) grants AST-1010004 and 
AST-1412981. The authors gratefully acknowledge the support which enabled these observations at 
the Keck, Gemini, Large Binocular Telescope, Very Large Telescope, Las Campanas, and Palomar 
Observatories. MWL thanks T.-K. Chan for a discussion on CGM simulations and Chi Po Choi for a 
discussion on statistics.
